---
title: "Introduction to Data Science - Midterm Project"
author: "Durjoy Roy (23-50576-1), Maruf Ahmmed (23-50604-1), Anuja Sarker (23-50624-1)"
date: "April 5, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. About the Dataset

The **Telco Customer Churn** dataset is sourced from the IBM Sample Dataset, 
available on Kaggle 
(https://www.kaggle.com/datasets/blastchar/telco-customer-churn). 
It contains information about a telecom company's customers, including 
demographic details, account information, and services subscribed. 
The target variable is **Churn** (Yes/No), indicating whether a customer 
left the company. The original dataset has 7043 rows and 21 variables. 
For this project, a random subsample of 1500 rows is used as per the 
course requirement.

# 2. Objectives

The objectives of this project are:

- To apply data preparation and cleaning techniques on a real-world dataset
- To handle missing values, outliers, duplicates, and inconsistent formats
- To perform Exploratory Data Analysis (EDA) on key variables
- To analyze and compare descriptive statistics across the two target 
  classes (Churn: Yes / No)
- To demonstrate normalization, encoding, filtering, balancing, and 
  train/test splitting techniques

# 3. Load Libraries and Dataset
```{r load-data}
library(tidyverse)
library(naniar)
library(ROSE)
library(caTools)
library(psych)
library(scales)
library(gridExtra)
library(VIM)

# Load dataset
raw_data <- read.csv("WA_Fn-UseC_-Telco-Customer-Churn.csv")

# Check dimensions
dim(raw_data)
head(raw_data, 5)
```

# 4. Subsample the Dataset
```{r subsample}
# Fix TotalCharges column (stored as character, should be numeric)
raw_data$TotalCharges <- as.numeric(raw_data$TotalCharges)

# Take random subsample of 1500 rows
set.seed(42)
df <- raw_data %>% sample_n(1500)

# Verify dimensions
dim(df)
cat("Rows:", nrow(df), "| Columns:", ncol(df))
```
**Interpretation:**
The original dataset contains 7043 rows. As per the course requirement 
of maximum 2000 rows, a random subsample of 1500 rows is taken using 
set.seed(42) to ensure reproducibility. TotalCharges was converted from 
character to numeric as it contained numeric values stored incorrectly.

# 5. Intentional Data Corruption

To simulate real-world data quality issues, the following problems 
are intentionally introduced into the dataset.
```{r corrupt-data}
df_dirty <- df

# 1. Introduce missing values in MonthlyCharges and tenure
set.seed(123)
na_idx1 <- sample(1:nrow(df_dirty), 60)
na_idx2 <- sample(1:nrow(df_dirty), 40)
df_dirty$MonthlyCharges[na_idx1] <- NA
df_dirty$tenure[na_idx2] <- NA

# 2. Introduce impossible/invalid values
df_dirty$tenure[5]  <- -99
df_dirty$tenure[10] <- 999
df_dirty$MonthlyCharges[15] <- -50

# 3. Introduce inconsistent format in gender column
df_dirty$gender[c(1,2,3)] <- "male"
df_dirty$gender[c(4,5)]   <- "FEMALE"

# 4. Add duplicate rows
dup_rows <- df_dirty[sample(1:nrow(df_dirty), 30), ]
df_dirty <- rbind(df_dirty, dup_rows)

# Verify corruption
cat("Total rows after duplicates:", nrow(df_dirty), "\n")
cat("Missing values - MonthlyCharges:", sum(is.na(df_dirty$MonthlyCharges)), "\n")
cat("Missing values - tenure:", sum(is.na(df_dirty$tenure)), "\n")
cat("Unique gender values:", unique(df_dirty$gender), "\n")
cat("Dataset saved as df_dirty for cleaning phase.\n")
```
**Interpretation:**
Four types of data issues are intentionally introduced:
1. Missing values: 60 in MonthlyCharges and 40 in tenure
2. Invalid values: negative and impossibly large values in tenure 
   and MonthlyCharges
3. Inconsistent format: gender column contains mixed case 
   values (male, FEMALE, Male, Female)
4. Duplicate rows: 30 duplicate rows added, bringing total to 1530

# 6. Data Cleaning

## 6.1 Visualize Missing Values
```{r missing-viz}
# Visualize missing values
vis_miss(df_dirty)
```
**Interpretation:**
The missing value plot shows that MonthlyCharges has approximately 
4% missing values and tenure has around 2.7% missing values. 
All other columns have no missing values. This visualization 
helps identify which variables require imputation before analysis.

## 6.2 Handle Missing Values
```{r missing-impute}
df_clean <- df_dirty

# Also impute TotalCharges NA if any remain
df_clean$TotalCharges[is.na(df_clean$TotalCharges)] <-
  median(df_clean$TotalCharges, na.rm = TRUE)

# Impute MonthlyCharges with median
df_clean$MonthlyCharges[is.na(df_clean$MonthlyCharges)] <- 
  median(df_clean$MonthlyCharges, na.rm = TRUE)

# Impute tenure with median
df_clean$tenure[is.na(df_clean$tenure)] <- 
  median(df_clean$tenure, na.rm = TRUE)

# Verify no missing values remain
cat("Missing in MonthlyCharges:", sum(is.na(df_clean$MonthlyCharges)), "\n")
cat("Missing in tenure:", sum(is.na(df_clean$tenure)), "\n")
```
**Interpretation:**
Missing values in MonthlyCharges and tenure are imputed using the 
median value of each respective column. Median is preferred over 
mean because it is robust to outliers, which is important since 
invalid values (like -99 and 999) are still present in the data 
at this stage. After imputation, both columns show 0 missing values.

## 6.3 Remove Invalid Values
```{r invalid-values}
# Check invalid values before fixing
cat("Invalid tenure values (before):\n")
cat("tenure = -99:", sum(df_clean$tenure == -99), "\n")
cat("tenure = 999:", sum(df_clean$tenure == 999), "\n")
cat("MonthlyCharges < 0:", sum(df_clean$MonthlyCharges < 0), "\n")

# Replace invalid values with median
valid_tenure_median <- median(
  df_clean$tenure[df_clean$tenure > 0 & df_clean$tenure <= 72], 
  na.rm = TRUE
)
valid_monthly_median <- median(
  df_clean$MonthlyCharges[df_clean$MonthlyCharges > 0], 
  na.rm = TRUE
)

df_clean$tenure[df_clean$tenure == -99] <- valid_tenure_median
df_clean$tenure[df_clean$tenure == 999] <- valid_tenure_median
df_clean$MonthlyCharges[df_clean$MonthlyCharges < 0] <- valid_monthly_median

# Verify
cat("\nInvalid tenure values (after):\n")
cat("tenure = -99:", sum(df_clean$tenure == -99), "\n")
cat("tenure = 999:", sum(df_clean$tenure == 999), "\n")
cat("MonthlyCharges < 0:", sum(df_clean$MonthlyCharges < 0), "\n")
```
**Interpretation:**
Invalid values were detected in tenure (-99 and 999) and 
MonthlyCharges (negative values), which are logically impossible 
for these variables. These are replaced with the median of valid 
values only (tenure between 0-72 months, MonthlyCharges above 0). 
After correction, all three checks return 0, confirming successful 
removal of invalid entries.

## 6.4 Fix Inconsistent Format
```{r fix-format}
# Check before fixing
cat("Gender values before fix:\n")
print(table(df_clean$gender))

# Standardize to title case
df_clean$gender <- str_to_title(df_clean$gender)

# Verify
cat("\nGender values after fix:\n")
print(table(df_clean$gender))
```
**Interpretation:**
The gender column contained inconsistent capitalization (male, 
FEMALE, Male, Female). Using str_to_title(), all values are 
standardized to title case (Female, Male). After fixing, only 
two unique values remain, confirming consistent formatting.

## 6.5 Remove Duplicate Rows
```{r remove-duplicates}
# Check duplicates before
cat("Total rows before removing duplicates:", nrow(df_clean), "\n")
cat("Duplicate rows found:", sum(duplicated(df_clean)), "\n")

# Remove duplicates
df_clean <- df_clean %>% distinct()

# Verify
cat("\nTotal rows after removing duplicates:", nrow(df_clean), "\n")
cat("Duplicate rows remaining:", sum(duplicated(df_clean)), "\n")
```
**Interpretation:**
30 duplicate rows were intentionally added in the corruption phase. 
Using distinct() from the dplyr package, all duplicate rows are 
identified and removed. The dataset returns to 1500 rows with 
0 duplicates remaining, confirming successful deduplication.

## 6.6 Outlier Detection
```{r outlier-detect}
# Boxplot to visualize outliers in numeric columns
p1 <- ggplot(df_clean, aes(y = MonthlyCharges)) +
  geom_boxplot(fill = "#3B8BD4", alpha = 0.7) +
  labs(title = "MonthlyCharges", y = "Value") +
  theme_minimal()

p2 <- ggplot(df_clean, aes(y = tenure)) +
  geom_boxplot(fill = "#1D9E75", alpha = 0.7) +
  labs(title = "Tenure", y = "Value") +
  theme_minimal()

p3 <- ggplot(df_clean, aes(y = TotalCharges)) +
  geom_boxplot(fill = "#EF9F27", alpha = 0.7) +
  labs(title = "TotalCharges", y = "Value") +
  theme_minimal()

grid.arrange(p1, p2, p3, ncol = 3)

# IQR method to count outliers
count_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_val <- Q3 - Q1
  sum(x < (Q1 - 1.5 * IQR_val) | x > (Q3 + 1.5 * IQR_val), na.rm = TRUE)
}

cat("Outliers in MonthlyCharges:", count_outliers(df_clean$MonthlyCharges), "\n")
cat("Outliers in tenure:", count_outliers(df_clean$tenure), "\n")
cat("Outliers in TotalCharges:", count_outliers(df_clean$TotalCharges), "\n")
```
**Interpretation:**
Boxplots are used to visually identify outliers in the three main 
numeric columns. The IQR (Interquartile Range) method is applied 
to count outliers mathematically — any value below Q1 - 1.5*IQR 
or above Q3 + 1.5*IQR is considered an outlier. This provides 
both a visual and quantitative understanding of extreme values 
in the dataset.

## 6.7 Handle Outliers
```{r outlier-handle}
# Cap outliers using IQR method (Winsorization)
cap_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_val <- Q3 - Q1
  lower <- Q1 - 1.5 * IQR_val
  upper <- Q3 + 1.5 * IQR_val
  x[x < lower] <- lower
  x[x > upper] <- upper
  return(x)
}

df_clean$MonthlyCharges <- cap_outliers(df_clean$MonthlyCharges)
df_clean$tenure <- cap_outliers(df_clean$tenure)
df_clean$TotalCharges <- cap_outliers(df_clean$TotalCharges)

# Verify
cat("Outliers after capping - MonthlyCharges:", 
    count_outliers(df_clean$MonthlyCharges), "\n")
cat("Outliers after capping - tenure:", 
    count_outliers(df_clean$tenure), "\n")
cat("Outliers after capping - TotalCharges:", 
    count_outliers(df_clean$TotalCharges), "\n")
```
**Interpretation:**
Outliers are handled using Winsorization (capping), where values 
below the lower bound (Q1 - 1.5*IQR) are replaced with the lower 
bound, and values above the upper bound (Q3 + 1.5*IQR) are replaced 
with the upper bound. This approach retains all rows while limiting 
the effect of extreme values. After capping, all three columns 
show 0 outliers.

# 7. Feature Engineering

## 7.1 Categorical to Numeric Encoding
```{r encoding}
# Check current data types
cat("Before encoding:\n")
cat("Churn:", class(df_clean$Churn), "| Values:", 
    unique(df_clean$Churn), "\n")
cat("gender:", class(df_clean$gender), "| Values:", 
    unique(df_clean$gender), "\n")
cat("Partner:", class(df_clean$Partner), "| Values:", 
    unique(df_clean$Partner), "\n")

# Convert binary categorical columns to numeric (0/1)
df_clean$Churn_numeric <- ifelse(df_clean$Churn == "Yes", 1, 0)
df_clean$gender_numeric <- ifelse(df_clean$gender == "Male", 1, 0)
df_clean$Partner_numeric <- ifelse(df_clean$Partner == "Yes", 1, 0)
df_clean$Dependents_numeric <- ifelse(df_clean$Dependents == "Yes", 1, 0)
df_clean$PhoneService_numeric <- ifelse(df_clean$PhoneService == "Yes", 1, 0)

# Verify
cat("\nAfter encoding:\n")
cat("Churn_numeric - unique values:", unique(df_clean$Churn_numeric), "\n")
cat("gender_numeric - unique values:", unique(df_clean$gender_numeric), "\n")
cat("Partner_numeric - unique values:", unique(df_clean$Partner_numeric), "\n")
```
**Interpretation:**
Five binary categorical columns (Churn, gender, Partner, Dependents, 
PhoneService) are converted to numeric format (0/1) using ifelse(). 
New columns are created with the suffix _numeric to preserve the 
original columns for reference. This encoding is necessary for 
mathematical operations and statistical analysis in later steps.

## 7.2 Numeric to Categorical Conversion
```{r numeric-to-cat}
# Convert tenure to age groups
df_clean$tenure_group <- cut(
  df_clean$tenure,
  breaks = c(0, 12, 36, 72),
  labels = c("New (0-12m)", "Mid (13-36m)", "Long (37-72m)"),
  include.lowest = TRUE
)

# Verify
cat("Tenure group distribution:\n")
print(table(df_clean$tenure_group))
```
**Interpretation:**
The continuous tenure column is converted into a categorical variable 
with three meaningful groups: New customers (0-12 months), Mid-term 
customers (13-36 months), and Long-term customers (37-72 months). 
This conversion allows group-based analysis and comparison, which 
is useful in the EDA phase.

## 7.3 Normalization

```{r normalize}
# Check range before normalization
cat("Before normalization:\n")
cat("MonthlyCharges - Min:", min(df_clean$MonthlyCharges, na.rm = TRUE), 
    "Max:", max(df_clean$MonthlyCharges, na.rm = TRUE), "\n")
cat("TotalCharges - Min:", min(df_clean$TotalCharges, na.rm = TRUE), 
    "Max:", max(df_clean$TotalCharges, na.rm = TRUE), "\n")
cat("tenure - Min:", min(df_clean$tenure, na.rm = TRUE), 
    "Max:", max(df_clean$tenure, na.rm = TRUE), "\n")

# Min-Max Normalization (Fixed with na.rm = TRUE)
min_max_norm <- function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}

df_clean$MonthlyCharges_norm <- min_max_norm(df_clean$MonthlyCharges)
df_clean$TotalCharges_norm <- min_max_norm(df_clean$TotalCharges)
df_clean$tenure_norm <- min_max_norm(df_clean$tenure)

# Verify
cat("\nAfter normalization:\n")
cat("MonthlyCharges_norm - Min:", round(min(df_clean$MonthlyCharges_norm, na.rm = TRUE), 4),
    "Max:", round(max(df_clean$MonthlyCharges_norm, na.rm = TRUE), 4), "\n")
cat("TotalCharges_norm - Min:", round(min(df_clean$TotalCharges_norm, na.rm = TRUE), 4),
    "Max:", round(max(df_clean$TotalCharges_norm, na.rm = TRUE), 4), "\n")
cat("tenure_norm - Min:", round(min(df_clean$tenure_norm, na.rm = TRUE), 4),
    "Max:", round(max(df_clean$tenure_norm, na.rm = TRUE), 4), "\n")
```
**Interpretation:**
Min-Max normalization is applied to MonthlyCharges, TotalCharges, 
and tenure, scaling all values to a range of 0 to 1. This ensures 
that no single variable dominates the analysis due to its larger 
scale. After normalization, all three columns correctly show 
a minimum of 0 and maximum of 1.

# 8. Data Filtering
```{r filtering}
# Filter 1: Senior Citizens who churned
senior_churn <- df_clean %>%
  filter(SeniorCitizen == 1 & Churn == "Yes")

cat("Senior citizens who churned:", nrow(senior_churn), "\n")

# Filter 2: Month-to-month contract customers
monthly_contract <- df_clean %>%
  filter(Contract == "Month-to-month")

cat("Month-to-month contract customers:", nrow(monthly_contract), "\n")

# Filter 3: High value customers (MonthlyCharges above average)
avg_charge <- mean(df_clean$MonthlyCharges, na.rm = TRUE)
high_value <- df_clean %>%
  filter(MonthlyCharges > avg_charge)

cat("High value customers (above avg charge):", nrow(high_value), "\n")
cat("Average MonthlyCharges:", round(avg_charge, 2), "\n")

# Filter 4: Long-term customers with no churn
loyal <- df_clean %>%
  filter(tenure > 36 & Churn == "No")

cat("Long-term loyal customers (tenure > 36m):", nrow(loyal), "\n")
```
**Interpretation:**
Four meaningful filters are applied to explore specific customer 
segments:
1. Senior citizens who churned — identifies vulnerable high-risk group
2. Month-to-month contract customers — this group is known to have 
   higher churn rates
3. High value customers — those paying above average monthly charges
4. Long-term loyal customers — tenure above 36 months with no churn

These filters help isolate specific subgroups for targeted analysis 
and support the findings section of the report.

# 9. Class Balancing
```{r class-balance}
# Check current class distribution
cat("Class distribution before balancing:\n")
print(table(df_clean$Churn))
cat("\nClass percentage:\n")
print(round(prop.table(table(df_clean$Churn)) * 100, 2))

# Visualize imbalance
ggplot(df_clean, aes(x = Churn, fill = Churn)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  scale_fill_manual(values = c("#3B8BD4", "#E8593C")) +
  labs(title = "Class Distribution Before Balancing",
       x = "Churn", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```
**Interpretation:**
The class distribution shows a significant imbalance — approximately 
76% customers did not churn (No) and 24% churned (Yes). This kind 
of imbalance can bias any future analysis or modeling, as the 
majority class dominates. A bar plot is used to visually confirm 
this imbalance before applying any balancing technique.

```{r balancing}
# Prepare clean dataframe for ROSE
df_rose <- df_clean %>%
  select(-customerID) %>% # শুধু ID বাদ দিচ্ছি, বাকি সব কলাম থাকবে
  mutate(
    TotalCharges = ifelse(is.na(TotalCharges), 
                          median(TotalCharges, na.rm = TRUE), 
                          TotalCharges)
  ) %>%
  mutate_if(is.character, as.factor) # সব character কে একবারে factor বানিয়ে দেওয়া

# Apply ROSE
set.seed(42)
df_balanced <- ROSE(Churn ~ ., data = df_rose, seed = 42)$data

# Check distribution after balancing
cat("Class distribution after balancing:\n")
print(table(df_balanced$Churn))
cat("\nClass percentage after balancing:\n")
print(round(prop.table(table(df_balanced$Churn)) * 100, 2))

# Visualize balanced classes
ggplot(df_balanced, aes(x = Churn, fill = Churn)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  scale_fill_manual(values = c("#3B8BD4", "#E8593C")) +
  labs(title = "Class Distribution After Balancing",
       x = "Churn", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```
**Interpretation:**
Before balancing, the dataset was prepared by handling missing values (using median imputation for TotalCharges) and converting character variables into factors to ensure compatibility with the algorithm. Then, ROSE (Random Over-Sampling Examples) is applied to generate a balanced dataset where both classes (Churn: Yes / No) have approximately equal representation. Unlike simple duplication, ROSE generates synthetic samples, making the balance more robust. After balancing, both classes show approximately 50% distribution, successfully eliminating the previous bias.

# 10. Train/Test Split
```{r train-test-split}
# Use df_clean for splitting (retains all original features)
set.seed(42)

# 80/20 split
split <- sample.split(df_clean$Churn, SplitRatio = 0.80)

train_data <- subset(df_clean, split == TRUE)
test_data  <- subset(df_clean, split == FALSE)

# Verify split
cat("Total rows in df_clean:", nrow(df_clean), "\n")
cat("Training set rows:", nrow(train_data), "\n")
cat("Testing set rows:", nrow(test_data), "\n")
cat("Train ratio:", round(nrow(train_data)/nrow(df_clean)*100, 2), "%\n")
cat("Test ratio:", round(nrow(test_data)/nrow(df_clean)*100, 2), "%\n")

# Verify class distribution is maintained in both sets
cat("\nChurn distribution in Training set:\n")
print(round(prop.table(table(train_data$Churn))*100, 2))

cat("\nChurn distribution in Testing set:\n")
print(round(prop.table(table(test_data$Churn))*100, 2))
```
**Interpretation:**
The cleaned dataset (df_clean) is split into training (80%) and 
testing (20%) sets using the caTools package. set.seed(42) ensures 
reproducibility. sample.split() is used to maintain the original 
class distribution (stratified split) in both sets, meaning the 
Churn ratio remains consistent across train and test data. 
The training set will be used to build models in future analysis, 
while the testing set evaluates model performance on unseen data.

# 11. Exploratory Data Analysis (EDA)

## 11.1 Descriptive Statistics across Two Target Classes

```{r descriptive-stats}
# Overall descriptive statistics
cat("=== Overall Descriptive Statistics ===\n")
df_clean %>%
  select(tenure, MonthlyCharges, TotalCharges) %>%
  summary() %>%
  print()

# Descriptive statistics by Churn class
cat("\n=== Statistics for Churn = No ===\n")
df_clean %>%
  filter(Churn == "No") %>%
  select(tenure, MonthlyCharges, TotalCharges) %>%
  summary() %>%
  print()

cat("\n=== Statistics for Churn = Yes ===\n")
df_clean %>%
  filter(Churn == "Yes") %>%
  select(tenure, MonthlyCharges, TotalCharges) %>%
  summary() %>%
  print()

# Standard deviation by class
cat("\n=== Standard Deviation by Churn Class ===\n")
df_clean %>%
  group_by(Churn) %>%
  summarise(
    SD_tenure = round(sd(tenure, na.rm = TRUE), 2),
    SD_MonthlyCharges = round(sd(MonthlyCharges, na.rm = TRUE), 2),
    SD_TotalCharges = round(sd(TotalCharges, na.rm = TRUE), 2)
  ) %>%
  print()
```
**Interpretation:**
Descriptive statistics are calculated separately for churned (Yes) 
and non-churned (No) customers. Key observations:
- Churned customers have significantly lower median tenure, 
  indicating they leave early in their subscription.
- MonthlyCharges are higher for churned customers on average, 
  suggesting price sensitivity as a churn driver.
- TotalCharges are lower for churned customers due to shorter tenure.
- Standard deviation of tenure is higher in the No class, meaning 
  loyal customers have more varied subscription lengths.
  
## 11.2 Average MonthlyCharges Comparison between Two Groups

```{r group-comparison}
# Calculate average MonthlyCharges by Churn group
avg_by_churn <- df_clean %>%
  group_by(Churn) %>%
  summarise(
    Mean_MonthlyCharges = round(mean(MonthlyCharges, na.rm = TRUE), 2),
    Median_MonthlyCharges = round(median(MonthlyCharges, na.rm = TRUE), 2),
    Mean_tenure = round(mean(tenure, na.rm = TRUE), 2),
    Median_tenure = round(median(tenure, na.rm = TRUE), 2),
    Count = n()
  )

cat("=== Average Comparison by Churn Class ===\n")
print(avg_by_churn)

# Visualize MonthlyCharges comparison
p1 <- ggplot(df_clean, aes(x = Churn, y = MonthlyCharges, fill = Churn)) +
  geom_bar(stat = "summary", fun = "mean") +
  scale_fill_manual(values = c("#3B8BD4", "#E8593C")) +
  labs(title = "Average MonthlyCharges by Churn",
       x = "Churn", y = "Average Monthly Charges") +
  theme_minimal() +
  theme(legend.position = "none")

# Visualize tenure comparison
p2 <- ggplot(df_clean, aes(x = Churn, y = tenure, fill = Churn)) +
  geom_bar(stat = "summary", fun = "mean") +
  scale_fill_manual(values = c("#3B8BD4", "#E8593C")) +
  labs(title = "Average Tenure by Churn",
       x = "Churn", y = "Average Tenure (months)") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p1, p2, ncol = 2)
```
**Interpretation:**
The average MonthlyCharges and tenure are compared across the two 
Churn classes. Key findings:
- Churned customers (Yes) pay higher average MonthlyCharges 
  compared to retained customers (No), suggesting that higher 
  billing is associated with increased churn likelihood.
- Retained customers (No) have significantly higher average tenure, 
  confirming that long-term customers are more loyal and less 
  likely to leave.
- These differences highlight MonthlyCharges and tenure as 
  important indicators of customer churn behavior.
  
## 11.3 Variance of MonthlyCharges across Contract Types

```{r variance-analysis}
# Variance and SD of MonthlyCharges across Contract types
variance_table <- df_clean %>%
  group_by(Contract) %>%
  summarise(
    Count = n(),
    Mean = round(mean(MonthlyCharges, na.rm = TRUE), 2),
    SD = round(sd(MonthlyCharges, na.rm = TRUE), 2),
    Variance = round(var(MonthlyCharges, na.rm = TRUE), 2),
    Min = round(min(MonthlyCharges, na.rm = TRUE), 2),
    Max = round(max(MonthlyCharges, na.rm = TRUE), 2)
  )

cat("=== Variance of MonthlyCharges by Contract Type ===\n")
print(variance_table)

# Boxplot to visualize spread
ggplot(df_clean, aes(x = Contract, y = MonthlyCharges, fill = Contract)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#3B8BD4", "#1D9E75", "#E8593C")) +
  labs(title = "Spread of MonthlyCharges across Contract Types",
       x = "Contract Type", y = "Monthly Charges") +
  theme_minimal() +
  theme(legend.position = "none")
```
**Interpretation:**
The variance and spread of MonthlyCharges are analyzed across 
three contract types. Key findings:
- Two-year contracts show the highest variance (1134) and the 
  widest spread in the boxplot. This indicates that long-term 
  customers choose a wide variety of plans, ranging from basic 
  to premium, resulting in greater price diversity.
- One-year contracts have moderate variance (1034), showing 
  intermediate price spread.
- Month-to-month contracts have the lowest variance (702), 
  suggesting that short-term customers tend to cluster around 
  a more consistent price range.
- The boxplot confirms this pattern — Two-year and One-year 
  contract boxes are visibly wider than Month-to-month.
  
## 11.4 Churn Rate by Contract Type

```{r churn-by-contract}
# Churn rate by contract type
churn_contract <- df_clean %>%
  group_by(Contract, Churn) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Contract) %>%
  mutate(Percentage = round(Count/sum(Count)*100, 2))

cat("=== Churn Rate by Contract Type ===\n")
print(churn_contract)

# Visualize
ggplot(churn_contract, aes(x = Contract, y = Percentage, fill = Churn)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("#3B8BD4", "#E8593C")) +
  geom_text(aes(label = paste0(Percentage, "%")),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  labs(title = "Churn Rate by Contract Type",
       x = "Contract Type", y = "Percentage (%)") +
  theme_minimal()
```
**Interpretation:**
Churn rate is analyzed across three contract types to identify 
which contract segment is most at risk. Key findings:
- Month-to-month contract customers have the highest churn rate, 
  confirming that short-term contracts are strongly associated 
  with customer attrition.
- Two-year contract customers show the lowest churn rate, 
  indicating that longer commitments lead to greater customer 
  retention.
- This finding directly supports the filtering result from 
  Section 8, where month-to-month customers were identified 
  as a high-risk segment.
  
# 12. Objectives and Findings Summary

## 12.1 Objectives Achieved

The following objectives set at the beginning of this project 
have been successfully completed:

- Data preparation and cleaning techniques were applied to a 
  real-world dataset containing intentionally introduced issues.
- Missing values, outliers, duplicates, and inconsistent formats 
  were detected and handled systematically.
- Exploratory Data Analysis was performed on key variables to 
  extract meaningful insights.
- Descriptive statistics were calculated and compared across 
  the two target classes (Churn: Yes / No).
- Normalization, encoding, filtering, class balancing, and 
  train/test splitting were demonstrated successfully.

## 12.2 Key Findings

**Finding 1 — Tenure and Churn:**
Churned customers have a significantly lower average tenure 
(17.8 months) compared to retained customers (38.1 months). 
Customers who leave do so early in their subscription period.

**Finding 2 — MonthlyCharges and Churn:**
Churned customers pay higher average MonthlyCharges (73.6) 
compared to retained customers (60.2). Higher billing is 
strongly associated with increased likelihood of churn.

**Finding 3 — Contract Type and Churn:**
Month-to-month contract customers show the highest churn rate 
(40.49%), while two-year contract customers show the lowest 
(1.78%). Longer contract commitments strongly reduce churn risk.

**Finding 4 — Variance across Contract Types:**
Two-year contracts show the highest variance in MonthlyCharges 
(1134), indicating diverse pricing plans among long-term 
customers. Month-to-month contracts have the lowest variance 
(702), suggesting more uniform pricing for short-term customers.

**Finding 5 — Class Imbalance:**
The dataset showed a significant class imbalance (76% No, 24% 
Yes). ROSE balancing successfully brought both classes to 
approximately equal distribution (52% vs 48%), making the 
data suitable for future modeling.

**Finding 6 — Senior Citizens:**
Filtered analysis identified senior citizens as a notable 
high-risk segment, with a higher proportion appearing among 
churned customers compared to retained customers.